---
title: "Week1_Reading Data"
author: "Bangda Sun"
date: "October 9, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### 1. Reading from MySQL
Free and widely used open source database; widely used in internet based app; data structured in: databases, table within databases, fields(column?) within tables. Each row is called a record.

example: http://dev.mysql.com/doc/employee/en/sakila-structure.html

#### Step 1-Install MySQL
http://dev.mysql.com/doc/refman/5.7/en/installing.html

#### Step 2-Install RMySQL
On windows: official instructions-http://biostat.mc.vanderbilt.edu/wiki/Main/RMySQL
Potentially useful guide: http://www.ahschulz.de/2013/07/23/installing-rmysql-under-windows/


#### Example - UCSC database
http://genome.ucsc.edu/
http://genome.ucsc.edu/goldenPath/help/mysql.html

```{r}
library(RMySQL)
# open connection
ucscDb <- dbConnect(MySQL(), user = "genome",
                    host = "genome-mysql.cse.ucsc.edu")
# query: show databases is a mysql command
result <- dbGetQuery(ucscDb, "show databases;");
# important: disconnect 
dbDisconnect(ucscDb);
result[1:10,]
```

#### Connecting to hg19 (database name) and listing table
```{r}
hg19 <- dbConnect(MySQL(), user = "genome", db = "hg19",
                  host = "genome-mysql.cse.ucsc.edu")
allTables <- dbListTables(hg19)
length(allTables)
# first 10 tables
allTables[1:10]
```

#### Get dimensions of a specific table
```{r}
dbListFields(hg19, "affyU133Plus2") # list fields in table: affyU133Plus2
# how many different rows/records in the dataset?
dbGetQuery(hg19, "select count(*) from affyU133Plus2")
```

#### Read from the table
```{r, warning = FALSE}
affyData <- dbReadTable(hg19, "affyU133Plus2")
head(affyData)
```

#### Select a specific subset
```{r, warning = FALSE}
query <- dbSendQuery(hg19, "select * from affyU133Plus2 where misMatches between 1 and 3")
affyMis <- fetch(query); quantile(affyMis$Matches)
# top 10 records, get a smmall dataset. and clear the query
affyMisSmall <- fetch(query, n = 10); dbClearResult(query)
dim(affyMisSmall)
```

#### Don't forget to close the connection!
```{r}
dbDisconnect(hg19)
```

#### Further resources
RMySQL vignette: http://cran.r-project.org/web/packages/RMySQL/RMySQL.pdf
List of commands: http://www.pantz.org/software/mysql/mysqlcommands.html. Do not, do not, delete, add or join things from ensemble. Only select. In general be careful with mySQL commands
A nice blog post summarizing some other commands: http://www.r-bloggers.com/mysql-and-r/

### 2. Reading from HDF5
Used for storing large data sets; support storing a range of data types; heirarchical data format; groups containing zero or more data sets and metadata(have a group header with group name and list of attributes; have a group symbol table with a list of objects in group); datasets multidimensional array of data elements with metadata(have a header with name, datatype, dataspace, and storage layout; have a data array with the data)
http://www.hdfgroup.org/

#### R HDF5 package
```
source("http://bioconductor.org/biocLite.R")
# rhdf5 pacakge
biocLite("rhdf5")
```
```{r}
library(rhdf5)
created = h5createFile("example.h5")
created
```
This whill install packages from Bioconductor http://bioconductor.org/, primarily used genomics but also has good "big data" packages.
Can be used to interface with hdf5 data sets.

#### Create groups
```{r}
created <- h5createGroup("example.h5", "foo")
created <- h5createGroup("example.h5", "baa")
created <- h5createGroup("example.h5", "foo/foobaa")
h5ls("example.h5")
```

#### Write to groups
```{r}
A <- matrix(1:10, nrow = 5, ncol = 2)
h5write(A, "example.h5", "foo/A")
B = array(seq(0.1, 2.0, by = .1), dim = c(5, 2, 2))
attr(B, "scale") <- "liter"
h5write(B, "example.h5", "foo/foobaa/B")
h5ls("example.h5")
```

#### Write a data set
```{r}
#df <- data.frame(1L:5L, seq(0, 1, length.out = 5), c("ab", "cde", "fghi", "a", "s"), stringsAsFactors = FALSE)
#h5write(df, "example.h5", "df")
#h5ls("example.h5")
```

#### Reading data
```{r}
readA  <- h5read("example.h5", "foo/A")
readB  <- h5read("example.h5", "foo/foobaa/B")
readdf <- h5read("example.h5", "df")
readA
readB
readdf
```

#### Writing and reading chunks
```{r}
# write into A[1:3, 1]
h5write(c(12, 13, 14), "example.h5", "foo/A", index = list(1:3, 1))
h5read("example.h5", "foo/A")
```

#### Notes
hdf5 can be used to optimize reading/writing from disc in R.

### 3. Reading data from the web
extracting data from the HTML code of website.
#### Example
http://scholar.google.com/citations?user=HI-I6C0AAAAJ&hl=en

#### Getting data off webpages - readLines()
```{r}
con <- url("http://scholar.google.com/citations?user=HI-I6C0AAAAJ&hl=en")
htmlCode <- readLines(con)
close(con)
htmlCode
```

#### Parsing with XML
```{r}
library(XML)
url <- "http://scholar.google.com/citations?user=HI-I6C0AAAAJ&hl=en"
html <- htmlTreeParse(url, useInternalNodes = TRUE)
xpathSApply(html, "//title", xmlValue)
xpathSApply(html, "//td[@id='col-citedby']", xmlValue) # nothing comes out...
```

#### Get from the httr package
```{r}
library(httr)
# take the same url as before
html2 <- GET(url)
# extract the content as text string
content2 <- content(html2, as = "text")
# parse command
parsedHtml <- htmlParse(content2, asText = TRUE)
# extract the title
xpathSApply(parsedHtml, "//title", xmlValue)
```

#### Accessing websites with passwords
```{r}
pg2 <- GET("http://httpbin.org/basic-auth/user/passwd", authenticate("user", "passwd"))
pg2
names(pg2)
```

#### Using handles
```{r}
google <- handle("http://google.com")
# go get path
pg1 <- GET(handle = google, path = "/")
pg2 <- GET(handle = google, path = "search")
```

### 4. Reading from the APIs
Applications programming interfaces
http://dev.twitter.com/docs/api1/get/blocks/blocking

#### Creating an application
https://dev.twitter.com/apps)

#### Accessing Twitter from R
```{r}
# oauth_app() is in the library httr
myapp <- oauth_app("twitter",
                   key = "yourConsumerKeyHere", secret = "yourConsumeSecretHere")
# sign in using token
sig <- sign_oauth1.0(myapp, token = "yourTokenHere", token_secret = "yourTokenSecretHere")
# GET command, using a specific url. get json file
homeTL <- GET("https://api.twitter.com/1.1/statuses/home_timeline.json", sig)
```

#### Converting the json object
```{r}
json1 <- content(homeTL)
library(jsonlite)
json2 <- jsonlite::fromJSON(toJSON(json1))
json2 # wrong result here
```

#### How did I know what url to use?
https://dev.twitter.com/docs/api/1.1/get/search/tweets

### 5. Reading from Other Sources

In general the best way to find out if the R package exist is to google "data storagr mechanism R package", for example: "MySQL R package".

#### Interacting more directly with files
`file` - open a connection to a text file
`url` - open a connection to a url
`gzfile` - open a connection to a .gz file
`bzfile` - open a connection to a .bz file
?connections for more info, and REMEMBER TO CLOSE CONNECTIONS!!

#### Foreign package
Loads data from Minitab, S, SAS, SPSS, Stata, Systat. Basic function `read.foo()`; `read.arff()` for Weka; `read.dta()` for Stata; `read.mtp()` for Minitab; `read.octave()` for Octave; `read.spss()` for SPSS; `read.xport()` for SAS.

#### Examples of other database packages
RPostresSQL; RODBC; RMongo.

#### Reading GIS data
rdgal; rgeos; raster.

#### Reading music data
tuneR; seewave.


















