---
title: "Week4_Text_Data"
author: "Bangda Sun"
date: "November 8, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Editing Text Variables

#### Fixing character vectors - tolower(), toupper()
```{r}
setwd("C://Users//Bangda//Dropbox//Get and Cleaning data in R//Week 4")
fileUrl <- "https://data.baltimorecity.gov/api/views/dz54-2aru/rows.csv?accessType=DOWNLOAD"
download.file(fileUrl, destfile = "cameras.csv", method = "wininet")
cameraData <- read.csv("cameras.csv")
# column name of the data
names(cameraData)
# convert the name into all lower case: tolower
tolower(names(cameraData))
# convert the name into all upper case: toupper
toupper(names(cameraData))
```

#### Fixing character vectors - strsplit()
Good to automatically splitting variable names
```{r}
splitName <- strsplit(names(cameraData), "\\.")
splitName
splitName[[5]]
splitName[[6]]
```

#### Qucik aside - lists
```{r}
mylist <- list(letters = c("A","B","C"), numbers = 1:3, matrix(1:25, ncol = 5))
mylist
# list element, still list
mylist[1]
# original data type
mylist$letters
mylist[[1]]
```

#### Fixing character vectors - sapply()
Applies a function to each element in a vector or list
```{r}
splitName[[6]][1]
firstElement <- function(x){x[1]}
sapply(splitName, firstElement)
```

#### Peer review data
```{r}
fileUrl1 <- "https://dl.dropboxusercontent.com/u/7710864/data/reviews-apr29.csv"
fileUrl2 <- "https://dl.dropboxusercontent.com/u/7710864/data/solutions-apr29.csv"
download.file(fileUrl1, destfile = "reviews.csv", method = "wininet")
download.file(fileUrl2, destfile = "solutions.csv", method = "wininet")
reviews <- read.csv("reviews.csv")
solutions <- read.csv("solutions.csv")
head(reviews, 2)
head(solutions, 2)
```

#### Fixing character vector - sub()
Important params: pattern, replacement, x
```{r}
names(reviews)
sub("_", "", names(reviews))
```

#### Fixing character vectors - gsub()
Mutiple case
```{r}
testName <- "this_is_a_test"
# only the 1st one is replaced
sub("_", " ", testName)
gsub("_", " ", testName)
```

#### Finding values - grep(), grepl()
```{r}
# where does Alameda apper in the data set
grep("Alameda", cameraData$intersection)
# count the number
table(grepl("Alameda", cameraData$intersection))
# sample: Alameda doesn't appear
cameraData <- cameraData[!grepl("Alameda", cameraData$intersection), ]
# more on grep()
grep("Alameda", cameraData$intersection, value = TRUE)
grep("Jeff", cameraData$intersection)
length(grep("Alameda", cameraData$intersection))
```

#### More useful string functions
```{r}
library(stringr)
nchar("Jeffrey Leek")
substr("Jeffrey Leek", 1, 7)
paste("Jeffrey", "Leek")
paste("Jeffrey", "Leek", sep = "..")
paste0("Jeffrey", "Leek")
str_trim("Jeff        ")
```

#### Important points about text in data sets
Names of variable should be:
(a) all lower case when possible; 
(b) descriptive;
(c) not duplicated;
(d) not have underscore or dots or white spaces;

### Regular Expressions I
Combination of literals and metacharacters; to draw an analogy with natural language, think of lieral text forming the words of this language, and the metacharacters defining its grammar; regular expressions have a rich set of metacharacters.

We need a way to express: (a) whitespace word boundaries; (b) sets of literals; (c) the beginning and end of a line; (d) alternatives ("war" or "peace") Metacharacters to the rescue!

#### Metacharacters
Some metacharacter represent the start of a line.
```
^i think
```
will match
```
i think we all rule for participating
i think i have been outed
i think this will be quite fun actually
```

Dollar sign represents the end of a line
```
morning$
```
will match
```
well they had something this morning
then had to catch a tram home in the morning
dog obedience school in the morning
```
#### Character Classes with []
We can list a set of characters we will accept at a given point in the match
```
[Bb][Uu][Ss][Hh]
```
will match BUSH (either upper or lower case)
```
The democarts are playing, "Name the worst thing about Bush"
BBQ and bushwalking at Molonglo Gorge
```

```
^[Ii] am
```
will match
```
i am so happy
I am a student
```
Similarly you can specify a range of letters [a-z] or [a-zA-Z]; notice that the order doesn't matter 
```
^[0-9][a-zA-Z]
```
will match
```
5th Avenue
2nd half of the game
```

When used at the beginning of a character class, the "" is also a metacharacter and indicates matching characters NOT in the indicated class
```
[^?.]$
```
will match 
```
i like basketballs
6 and 9
Not in Baghdad
```

### Regular Expression II
More Metacharacters
"." is used to refer to any character. So
```
9.11
```
will match
```
its stupid the post 9-11 rules
Front Door 9:11:46 AM
```
"|" is the meaning of "OR"
```
flood|earthquake|coldfire
```
will match 
```
any sentence including flood, earthquake, coldfire
```
The alternatives can be real expression and not just literals
```
^[Gg]ood|[Bb]ad
```
will match
```
good to hear some good knews from someone here
Good afternoon fellow american infidels
Katie...guess they had bad expriences...
my middle name is trouble, Miss Bad News
```
Subexpressions are often contained in parentheses to constrain the alternatives
```
^([Gg]ood|[Bb]ad)
```
will match
```
bad habbit
bad idea
good idea
```
The question mark "?" indicates that the indicated expression is optional
```
[Gg]eorge([Ww]\.)?[Bb]ush
```
will match
```
i bet i can spell better than you and george bush combined
a bird in the hand is worth two george w. bushes
```
The "*" and "+" signs are metacharacters used to indicate repetition; * means "any number including none, of the item" and + means "at least one of the item".
```
(.*)
```
will match
```
()
(24, m)
```
```
[0-9]+(.*)[0-9]+
# at least one number + any characters + at least one number
```
{ and } are referred to as interval quantifiers; the let us specify the minimum and maximum number of matches of an expression
```
[Bb]ush( +[^ ]+ +){1,5} debate
```
will match 
```
Bush (at least one space), followed by something not a space, followed by at least one space, between 1 and 5 times.

```
```
{m,}
# at least m matches
```
( and ). In most implementations of regular expressions, the parentheses not only limit the scope of alternatives divided by a "|", but also can be used to "remember" text matched by the subexpression enclosed.
We refer to the matched text with \1, \2, etc.
```
 +([a-zA-Z]+) +\1 +
```
will match (word repetition more than 1 times)
```
time for bed, night night twitter!
blah blah blah
my tatoo is so so itchy today
hi anybody anybody at home
```
The greediness of * can be turned off with the ?, as in
```
^s(.*?)s$
```
Some functions: `grep()`, `grepl()`, `sub()`, `gsub()`.

### Working with dates
Starting simple
```{r}
d1 <- date(); d1
class(d1)
d2 <- Sys.Date(); d2
class(d2)
```

#### Formatting dates
%d = days as number (0-31), %a = abbreviated weekday, %A = unabbreviated weekday, %m = month (00-12), %b = abbreviated month, %B = unabbreviated month, %y = 2digit year, %Y = four digit year
```{r}
format(d2, "%a %b %d")
format(d2, "%A %B %d")
```

#### Creating dates
```{r}
x <- c("1jan1960", "2jan1960", "31mar1960", "30jul1960")
# turn into dates
z <- as.Date(x, "%d%b%Y")
z
z[1] - z[2] 
as.numeric(z[1] - z[2])
```

#### Converting to Julian
```{r}
weekdays(d2)
months(d2)
months(z)
# number of days have occurs after the origin
julian(d2)
julian(z)
```

#### Lubridate
```{r}
library(lubridate)
# year + month + day
ymd("20140108")
mdy("03/04/2013")
dmy("03-04-2013")
```

#### Dealing with times
```{r}
ymd_hms("2016-11-08 10:57:26")
ymd_hms("2016-11-08 10:57:26", tz = "Pacific/Auckland")
# ?Sys.timezone
```

#### Some functions have slightly different syntax
```{r}
# in Lubridate package
x <- dmy(c("1jan2013", "2jan2013", "31mar2013", "30jul2013"))
wday(x[1])
wday(x[1], label = TRUE)
```




























